Comparing Hydrogen and Jet-A for an N+3 Turbofan with Water Recirculation using Gradient-Based Optimization

### 2023-05-01 conference-paper/main.tex
Added conclusion, other minor changes

% Message 2: Background and references to support using H2 and water injection in HBTF engines
% [ ] TODO: AL-PA, Need to find a journal/conference paper to cite for the following statement.  Maybe find one or more papers other than just the NASA report?
Water injection is the process of injecting water upstream of the combustor as finely atomized droplets.
NASA, Boeing, and Rolls-Royce studied this concept and suggested that this technique reduces the NOx emissions as much as 47 percent~\cite{Daggett2010}.
Additionally, water injection improves fuel efficiency and thrust output with lower combustion temperatures that can improve the lifetime of turbine blades and reduce noise~\cite{Daggett2010}.
% [x] TODO: AL-PA, Should try to find a citation for this if one exists.
Traditional propulsion systems that burn hydrocarbon fuels would require external water storage on the aircraft for water injection~\cite{Mourouzidis2015}.
The added weight of tanks, pumping, and ducting makes this concept infeasible for a conventional aircraft.
The main product of hydrogen combustion is water vapor and can thus be recovered from the exhaust stream~\cite{Strom2002}.
Condensing water vapor from the exhaust stream of hydrogen combustion and recirculating it elimiates the requirement for storage tanks.
This allows for the theoretical design of a closed loop water feedback system inside the propulsion cycle.
Indeed, the aerospace industry is starting to look into this technology to make hydrogen-powered aircraft economical~\footnote{https://arpa-e.energy.gov/technologies/projects/hydrogen-steam-and-inter-cooled-turbine-engine-hysite}{}.

@@

% [x] TODO: AL - PA, Define the T4 ratio with an equation and reference the equation in the table
% [x] TODO: AL-PA, I thought we were only doing recovery fractions at CRZ??
% [x] TODO: AL-PA, Make sure you reference the equations for the DVs/cons in the descriptions so the reader can connect them.
\begin{table}[hbt!]
    \centering
    \caption{
        Multipoint optimization problem definition.
        The objective function is the thrust specific energy consumption at the cruise condition.
        The constraints are net thrust and engine diameter constraint at the design point, TOC.
    }
    \small
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{r l l r l}
        \toprule
                        & Variable/Function              & Description                                                           & Units          & Quantity \\
        \hline
        minimize        & $\rm{TSEC}_{\rm{CRZ}} $        & Thrust-specific energy consumption at CRZ (Equation \eqref{eq:tsec})  & \si{lbm/s}     & 1        \\
                        &                                &                                                                       &                &          \\
        with respect to & $X_{\rm{H_2O,CRZ}}$            & Water recovery fraction at CRZ                                        & -              & 1        \\
                        & $\rm{PR}_{\rm{fan,TOC}}$       & TOC fan pressure ratio                                                & -              & 1        \\
                        & $\rm{PR}_{\rm{LPC,TOC}}$       & TOC low-pressure compressor pressure ratio                            & -              & 1        \\
                        & $\rm{OPR}_{\rm{TOC}}$          & TOC overall pressure ratio                                            & -              & 1        \\
                        & $T_{4,\rm{RTO}}$               & RTO combustor temperature                                             & \si{\degree R} & 1        \\
                        & $T_{4,\rm{ratio}}$             & TOC-to-RTO temperature ratio (Equation \eqref{t4ratio})               & -              & 1        \\
                        & $V_{\rm{ratio,CRZ}}$           & Core-to-bypass nozzle velocity ratio at CRZ (Equation \eqref{vratio}) & -              & 1        \\
        \cline{3-5}
                        &                                & Total                                                                 &                & 7        \\
                        &                                &                                                                       &                &          \\
        subject to      & $F_{\rm{net,TOC}} \geq 5800.0$ & Target net thrust at TOC                                              & \si{lbf}       & 1        \\
                        & $D_{\rm{fan}} \leq 100$        & Maximum Fan Diameter                                                  & \si{inch^2}    & 1        \\
        \cline{3-5}
                        &                                & Total                                                                 &                & 2        \\
        \bottomrule
    \end{tabular}
    \label{tab:opt_problem}
\end{table}

@@

From this plot we can visually check that each optimization achieved optimality and feasibility, and that the merit function reached a steady-state value.
Indeed, we see quadratic convergence in the optimiality of the problem near the optimal solution as expected.
The design variable history for each optmization problem is shown in Figure \ref{fig:history_dvs}.

@@

% NOTE: PA - PA/AL replace this with a parallel axis plot?
% [x] TODO: AL-PA, Yeah let's do the parallel axis plot for the basic thermodynamic parameters
% AL-PA, We can make another plot for the water fractions and the dpQp study.
\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{N3_parallel_coords.pdf}
    \caption{Optimization results of the N+3 engine with and without no water recovery using Jet-A and H2 as the fuel.
        The red lines show Jet-A fuel and the blue lines show H2 fuel.
        The solid lines show the engine without water recovery and the dashed lines show the engine with water recovery.}
    \label{fig:parallel_coords}
\end{figure}

We can see that the resulting engine designs are all very similar.
The main differences in the engine designs are the pressure ratios and temperatures.
Between the two fuel types, the water recovery fraction for both is pushed up to the upper limit of the design space.
Given that more water is present in the exhaust stream of the H2 engine, the H2 engine can recover a smaller fraction of the water compared to the Jet-A engine.
The water recovery flow rates for the Jet-A and H2 engines was 0.324 $\si{lbm/s}$ and 0.405 $\si{lbm/s}$, respectively.
Therefore, the H2 has the benefit of being able to recover more water at a lower fraction of the overall core exhaust stream water.
With water recovery, we see slightly higher pressure ratios across the fan and overall pressure ratio while the LPC pressure ratio is pushed to its technological upperbound.
We only see differences in $T_{4,\rm{RTO}}$ without water recovery due to these values are fixed for water recovery.
The $T_{4,\rm{RTO}}$ value for Jet-A is slightly higher than that of H2.
Furthermore, adding water recovery slightly decreases the $T_{4,\rm{ratio}}$ value.
Therefore, temperatures at cooler with the H2 fuel and barely cooler when adding water recovery.
Both fuels have similar efficiencies at the optimal design with Jet-A having a slight advantage in thems of TSEC.
Given that most of the design variables change only slightly between designs, most of the efficiency gains are driven by water injection.
In fact, we can see the $\rm{TSEC}_{\rm{CRZ}}$ improvement due to water recovery relative to no water recovery for both fuels in Figure \ref{fig:barchart}.

@@

Based on this analysis and model, we see an improvement greater than 5\% at CRZ for TSEC for both fuels.
This level of improvement would have a significant reduction in fuel costs over the course of an engine's lifecycle.
Due to the comparatively small amount of water in the exhaust stream for Jet-A, the Jet-A engine would require a larger condenser~\cite{Strom2002}.
This would incure larger condenser mass and pressure penalties compared to the H2 engine which would have more water and requires a smaller fraction of that water.
Therefore, H2 is a much more appealing fuel to use with the proposed closed-loop water recovery system.
Since no large-scale water condensers have been designed to fit inside a turbofan engine, a relatively low pressure loss condenser would need to be developed for this system to work.
Given the lack of knowledge of the type of condenser that would need to be require, we performed a pressure loss sweep of the H2 engine.
pyCycle handles pressure losses in ducts by specifyign a pressure loss coefficient, dPqP.
This coefficient is the ratio of change in pressure normalized by the inlet pressure of the duct.
The pressure loss of the duct just upstream of the extractor was varied while the rest of the engine design variables were optimized.
Figure \ref{fig:dpqp_sweep} shows $\rm{TSEC}_{\rm{CRZ}}$ of this sweep relative to and engine with no water recovery.

@@

From this figure, we can see the design area in which the closed-loop water recovery model is beneficial.
This model will offer efficiency benefits up until about $\rm{dPqP} = 0.16$ which means a 16\% loss of pressure from the condenser.
Furthermore, Figure \ref{fig:dpqp_wdot} shows the total water available in the core stream as a function of the same pressure loss.

@@

The breakdown of thrust for each point is shown in Table \ref{tab:thrust}.
This table shows the resulting net thrust, gross thrust, and ram drag of each engine design for the four optimizations performed.

@@

% Motivation: space to work with hydrogen, benefits of water injection, benefits of H2 (a lot of water, no tanks, de-mineralized water)
\section{Conclusion}
\label{sec:conc}
% Message 1: Re-state the problem
% Message 2: Summarize the HBTF model
% Message 3: Expected results

>>>>

% Message 2: Background and references to support using H2 and water injection in HBTF engines
% [ ] TODO: AL-PA, Need to find a journal/conference paper to cite for the following statement.  Maybe find one or more papers other than just the NASA report?
Water injection is the process of injecting water upstream of the combustor as finely atomized droplets.
NASA, Boeing, and Rolls-Royce studied this concept and suggested that this technique reduces the NOx emissions as much as 47 percent~\cite{Daggett2010}.
Additionally, water injection improves fuel efficiency and thrust output with lower combustion temperatures that can improve the lifetime of turbine blades and reduce noise~\cite{Daggett2010}.
% [x] TODO: AL-PA, Should try to find a citation for this if one exists.
Traditional propulsion systems that burn hydrocarbon fuels would require external water storage on the aircraft for water injection~\cite{Mourouzidis2015}.
The added weight of tanks, pumping, and ducting makes this concept infeasible for a conventional aircraft.
The main product of hydrogen combustion is water vapor and can thus be recovered from the exhaust stream~\cite{Strom2002}.
Condensing water vapor from the exhaust stream of hydrogen combustion and recirculating it elimiates the requirement for storage tanks.
This allows for the theoretical design of a closed loop water feedback system inside the propulsion cycle.
Indeed, the aerospace industry is starting to look into this technology to make hydrogen-powered aircraft economical~\footnote{https://arpa-e.energy.gov/technologies/projects/hydrogen-steam-and-inter-cooled-turbine-engine-hysite}.

@@

% [x] TODO: AL - PA, Define the T4 ratio with an equation and reference the equation in the table
% [x] TODO: AL-PA, I thought we were only doing recovery fractions at CRZ??
% [x] TODO: AL-PA, Make sure you reference the equations for the DVs/cons in the descriptions so the reader can connect them.
% [ ] TODO: PA-PA, explain variable ranges
\begin{table}[hbt!]
    \centering
    \caption{
        Multipoint optimization problem definition.
        The objective function is the thrust specific energy consumption at the cruise condition.
        The constraints are net thrust and engine diameter constraint at the design point, TOC.
    }
    \small
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{r l l r l}
        \toprule
                        & Variable/Function              & Description                                                           & Units          & Quantity \\
        \hline
        minimize        & $\rm{TSEC}_{\rm{CRZ}} $        & Thrust-specific energy consumption at CRZ (Equation \eqref{eq:tsec})  & \si{lbm/s}     & 1        \\
                        &                                &                                                                       &                &          \\
        with respect to & $X_{\rm{H_2O,CRZ}}$            & Water recovery fraction at CRZ                                        & -              & 1        \\
                        & $\rm{PR}_{\rm{fan,TOC}}$       & TOC fan pressure ratio                                                & -              & 1        \\
                        & $\rm{PR}_{\rm{LPC,TOC}}$       & TOC low-pressure compressor pressure ratio                            & -              & 1        \\
                        & $\rm{OPR}_{\rm{TOC}}$          & TOC overall pressure ratio                                            & -              & 1        \\
                        & $T_{4,\rm{RTO}}$               & RTO combustor temperature                                             & \si{\degree R} & 1        \\
                        & $T_{4,\rm{ratio}}$             & TOC-to-RTO temperature ratio (Equation \eqref{t4ratio})               & -              & 1        \\
                        & $V_{\rm{ratio,CRZ}}$           & Core-to-bypass nozzle velocity ratio at CRZ (Equation \eqref{vratio}) & -              & 1        \\
        \cline{3-5}
                        &                                & Total                                                                 &                & 7        \\
                        &                                &                                                                       &                &          \\
        subject to      & $F_{\rm{net,TOC}} \geq 5800.0$ & Target net thrust at TOC                                              & \si{lbf}       & 1        \\
                        & $D_{\rm{fan}} \leq 100$        & Maximum Fan Diameter                                                  & \si{inch^2}    & 1        \\
        \cline{3-5}
                        &                                & Total                                                                 &                & 2        \\
        \bottomrule
    \end{tabular}
    \label{tab:opt_problem}
\end{table}

\begin{table}[hbt!]
    \centering
    \caption{Design variable ranges for the optimization problem.
    }
    \small
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{l l l l}
        Design Variable       & Lower  & Upper  & Units     \\
        \toprule
        $x_{H2O,CRZ}$         & 0.0    & -      & -         \\
        $PR_{fan,TOC}$        & 1.2    & 1.4    & -         \\
        $PR_{LPC,TOC}$        & 2.5    & 4.0    & -         \\
        $PR_{OPR,TOC}$        & 40.0   & 70.0   & -         \\
        $T_{4,TOC}/T_{4,RTO}$ & 0.5    & 0.95   & -         \\
        $T_{4,RTO}$           & 3000.0 & 3600.0 & $^\circ$R \\
        $V_{ratio,CRZ}$       & 1.35   & 1.45   &           \\
        \bottomrule
    \end{tabular}
    \label{tab:res_full_opt_H2}
\end{table}

@@

From this plot we can visually check that each optimization achieved optimality and feasibility, and that the merit function reached a steady-state value.
Indeed, we observe quadratic convergence in the optimiality of the problem near the optimal solution as expected.
The design variable history for each optmization problem is shown in Figure \ref{fig:history_dvs}.

@@

% NOTE: PA-PA/AL replace this with a parallel axis plot?
% [x] TODO: AL-PA, Yeah let's do the parallel axis plot for the basic thermodynamic parameters
% AL-PA, We can make another plot for the water fractions and the dpQp study.
\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{N3_parallel_coords.pdf}
    \caption{Optimization results of the N+3 engine with and without no water recovery using Jet-A and H2 as the fuel.
        The red lines show Jet-A fuel and the blue lines show H2 fuel.
        The solid lines show the engine without water recovery and the dashed lines show the engine with water recovery.}
    \label{fig:parallel_coords}
\end{figure}

We can see that the resulting engine designs are all very similar.
The main differences in the engine designs are the pressure ratios and temperatures.
Between the two fuel types, the water recovery fraction for both is pushed up to the upper limit of the design space.
Given that more water is present in the exhaust stream of the H2 engine, the H2 engine can recover a smaller fraction of the water compared to the Jet-A engine.
The water recovery flow rates for the Jet-A and H2 engines was 0.324 $\si{lbm/s}$ and 0.405 $\si{lbm/s}$, respectively.
Therefore, the H2 has the benefit of being able to recover more water at a lower fraction of the overall core exhaust stream water.
With water recovery, we see slightly higher pressure ratios across the fan and overall pressure ratio while the LPC pressure ratio is pushed to its technological upperbound.
We only see differences in $T_{4,\rm{RTO}}$ without water recovery due to these values are fixed for water recovery.
The $T_{4,\rm{RTO}}$ value for Jet-A is slightly higher than that of H2.
Furthermore, adding water recovery slightly decreases the $T_{4,\rm{ratio}}$ value.
Therefore, temperatures are cooler with the H2 fuel and even slightly cooler when adding water recovery.
It is noted that with more water recovery the farther the optimizer would push the value of $T_{4,\rm{RTO}}$ down.
Therefore, there are temperature-related advantages of using water recovery to maintain a certain effenciency and thrust.
The breakdown of thrust for each point is shown in Table \ref{tab:thrust}.
This table shows the resulting net thrust, gross thrust, and ram drag of each engine design for the four optimizations performed.

From this table we can see that the net thrust is constrained at values set by \emph{balance} components or optimization constraints.
We see the largest thrust differences at the RTO and CRZ points.
Since net thrust is constrainted, the optimizer sizes the engine to reduce ram drag slightly from Jet-A to H2.
Water recirculation provides an even more significant improvement in ram drag for both Jet-A and H2 at CRZ with reductions of 0.21\% and 0.24\%, respectively.
Both fuels have similar efficiencies at the optimal design with Jet-A having a slight advantage in thems of TSEC.
Given that most of the design variables change only slightly between designs, most of the efficiency gains are driven by water injection.
In fact, we can see the $\rm{TSEC}_{\rm{CRZ}}$ improvement due to water recovery relative to no water recovery for both fuels in Figure \ref{fig:barchart}.

@@

Based on this analysis and model, we see an improvement greater than 5\% at CRZ for TSEC for both fuels.
This level of improvement would have a significant reduction in fuel costs over the course of an engine's lifecycle.
Due to the comparatively small amount of water in the exhaust stream for Jet-A, the Jet-A engine would require a larger condenser~\cite{Strom2002}.
This would incure larger condenser mass and pressure penalties compared to the H2 engine which would have more water and requires a smaller fraction of that water.
Therefore, H2 is a much more appealing fuel to use with the proposed closed-loop water recovery system.
Since there a lack of literature on large-scale water condensers that are designed to fit inside a turbofan engine, a relatively low pressure loss condenser would need to be developed for this system to work.
Given the lack of knowledge of the type of condenser that would need to be require, we performed a pressure loss sweep of the H2 engine.
pyCycle handles pressure losses in ducts by specifyign a pressure loss coefficient, dPqP.
This coefficient is the ratio of change in pressure normalized by the inlet pressure of the duct.
The pressure loss of the duct just upstream of the extractor was varied while the rest of the engine design variables were optimized.
Figure \ref{fig:dpqp_sweep} shows $\rm{TSEC}_{\rm{CRZ}}$ of this sweep relative to and engine with no water recovery.

@@

From this figure, we can see the design area in which the closed-loop water recovery model is beneficial.
This model will offer efficiency benefits up until about $\rm{dPqP} = 0.16$ which means a 16\% loss of pressure from the condenser.
This benefit area would be reduced if a smaller fraction of water recovered could be obtained.
Figure \ref{fig:dpqp_wdot} shows the total water available in the core stream as a function of the same pressure loss.

@@

@@

% Motivation: space to work with hydrogen, benefits of water injection, benefits of H2 (a lot of water, no tanks, de-mineralized water)
\section{Conclusion}
\label{sec:conc}
In this paper, we describe a how a novel closed-loop water recirculation system would be integrated in a commercial aviation setting.
The water recirculation system comprised was designed and implemented into the N+3 ultra-high bypass engine model.
We have presented the results of four optimization problems demonstrating the use of this technology using the pyCycle cycle modeling tool for Jet-A and hydrogen fuel.

The recirculation system is implemented by introducing two new custom pyCycle components, the water extractor and injector, which form a continuous closed-loop feedback water flow in the engine.
We then described the multipoint design problem and how it is used to design an aircraft engine to meet the operating requirements at different flight regimes.
System-level \emph{balance} constraints on the engine are described to show how a reduced-space optimization problem can be developed.
The optimization problem is then introduced with constraints on the design point thrust and overall engine size with thermodynamic design variables describing the temperatures, presusres and velocities within the engine.
An additional design variable controlling how much water is recirulated is added at the CRZ point to improve the fuel burn.

The results from the optimization problems show temperature, thrust, and efficiency improvements when using hydrogen as a fuel with water circulation.
Thermal reductions will improve the lifetime of the engine which impact maintanence and overall costs.
Between the two fuels, we see similar values of TSEC.
Significant energy efficiency gains are seen with water recirculation and injection.
This would lead to lower operating costs over the lifetime of the engine.

Technological problems do exist such as how to design a condensor that can recover a large portion of the exhaust water without large pressure losses.
However, the thermal properties of hydrogen provide a heat sink and thus a readily available resource for condensation.
Furthermore, using the exhaust stream to heat the hydrogen fuel has been proposed as an additional avenue to improve engine efficiency%~\cite{}.
The results from this work outline the working space available for the design of such a condenser.
% heavier engine due to condensor, but less fuel needed
